---
title: "04 Big Data Analysis"
author: "Darius Görgen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{04 Big Data Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
   %\VignetteDepends{ggplot2, plotly, stringr}
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F, cache = F)
```

```{r setup}
library(sf)
library(mapme.forest)
```


The previously demonstrated R routines base on the `raster` package which uses
quite a lot of RAM and this can be inefficient when it comes to the analysis of
very large rasters. 
In order to make the analysis of extensive regions feasible we developed a routine
using [GRASS GIS](https://grass.osgeo.org/) for the complete processing chain from
pre-processing GFW data to the extraction of zonal statistics.
This routine is very RAM efficient rendering the analysis of large areas possible
even when only restricted hardware is available (8GB of RAM should be enough for most
use cases). 

Equally to most of the other functions, it is expected that the path to the GDAL
binaries is exposed in your environment variable (refer to the README of this package
for more information). Additionally, the path to a working installation of GRASS GIS 
version >=7.0.0 is required for the function to work properly. Internally, the 
function uses the [`r.area`](https://grass.osgeo.org/grass78/manuals/addons/r.area.html) 
addon, which will be installed automatically if it is not present in the specified 
addon path option.

As inputs to the function it is required to specify the path to the tree cover,
loss year and CO2 emission files which cover the area of the polygons you are interested in.
We advise you to use the files you obtained by calling `downloadGFW()` with your 
area of interest, however, also raster files obtained in another way are suitable for 
the routine. 

The function call with the sample data from this package then looks like this:

```{r, eval = F}
# get the file paths to the raster files
treeCover = system.file("extdata", "pkgTest_treecover2000.tif", package = "mapme.forest")
lossYear = system.file("extdata", "pkgTest_lossyear.tif", package = "mapme.forest")
co2Layer = system.file("extdata", "pkgTest_co2_emission.tif", package = "mapme.forest")
# read the area of interest
roi = st_read(system.file("extdata", "aoi_polys.gpkg", package = "mapme.forest"))

grass = "/usr/lib/grass78"

roi_stats = statsGRASS(grass = grass, 
                       addon_base = "./data-raw/addons", 
                       areas = roi[1,], 
                       tree_cover = treeCover, 
                       tree_loss = lossYear, 
                       tree_co2 = co2Layer, 
                       idcol =  "id", 
                       thresholdClump = 10, 
                       thresholdCover = 50, 
                       years = 2001:2018, 
                       saveRaster = F)

```

This will deliver us the sf-object which entered the function with its dataframe
amended by the yearly forest cover area, tree cover loss area (both in ha) and 
CO2 emissions. Additionally, we can turn the `saveRaster` switch to TRUE and
specify a `outdir` variable in cases we want the raster output be written to disk.
The first layer then represents the raw GFW tree cover with values between 0 and 100,
the second layer the loss year layer, in this case with values between 0 and 18, 
the third layer the values of the CO2 emission equivalent, and then as many layers
as specified in the years options starting from the year 2000.

As mentioned above, this function runs very efficiently. You can hand it a very
high number of polygons simultaneously, each uniquely identified by the value in the
column you specify in `idcol`. Make sure the raster files you hand to the function
do cover all polygons and the function will clip and process for each polygon
after another. For the cases you selected to save the resulting raster files, 
inside the specified `outdir` you will find a `GTiff` for each polygon named 
according to the value in `idcol`.

