---
title: "03 Fragmentation Statistics"
author: "Darius Görgen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{03 Fragmentation Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
    %\VignetteDepends{ggplot2, plotly, stars, tmap, stringr}
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F, cache = F)
```

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning=F, message=F, cache = F)
library(raster)
library(stars)
library(igraph)
library(tibble)
library(magrittr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(sf)
library(plotly)
library(tmap)
library(mapme.forest)

```

## Fragmentation Indicators

Fragmentation statistics area calculated based on the package `SDMTools` which
itself is based on FRAGSTATS (McGarigal et al. 2012). We implemented support for
the following indicators:


**Available Fragmentation Statistics**

| **Name**                | **Explanation**                                              |
| ----------------------- | ------------------------------------------------------------ |
| n.patches               | the number of patches of a particular patch type or in a class. |
| total.area              | the sum of the areas (m2) of all patches of the corresponding patch type. |
| prop.landscape          | the proportion of the total lanscape represented by this class |
| patch.density           | the numbers of patches of the corresponding patch type divided by total landscape area (m2). |
| total.edge              | the total edge length of a particular patch type.            |
| edge.density            | edge length on a per unit area basis that facilitates comparison among landscapes of varying size. |
| mean.patch.area         | average area of patches.                                     |
| mean.frac.dim.index     | mean of fractal dimension index.                             |
| lanscape.division.index | based on the cumulative patch area distribution and is interpreted as the probability that two randomly chosen pixels in the landscape are not situated in the same patch |
| patch.cohesion.index    | measures the physical connectedness of the corresponding patch type. |
| landscape.shape.index   | a standardized measure of total edge or edge density that adjusts for the size of the landscape. |
| largest.patch.index     | largest patch index quantifies the percentage of total landscape area comprised by the largest patch. |
| sd.patch.area           | standard deviation of patch areas.                           |
| min.patch.area          | the minimum patch area of the total patch areas.             |
| max.patch.area          | the maximum patch area of the total patch areas.             |
| perimeter.area.frac.dim | perimeter-area fractal dimension equals 2 divided by the slope of regression line obtained by regressing the logarithm of patch area (m2) against the logarithm of patch perimeter (m). |
| mean.perim.area.ratio   | the mean of the ratio patch perimeter. The perimeter-area ratio is equal to the ratio of the patch perimeter (m) to area (m2). |
| sd.perim.area.ratio     | standard deviation of the ratio patch perimeter.             |
| max.perim.area.ratio    | maximum perimeter area ratio.                                |
| mean.shape.index        | mean of shape index                                          |
| sd.shape.index          | standard deviation of shape index.                           |
| min.shape.index         | the minimum shape index.                                     |
| max.shape.index         | the maximum shape index.                                     |
| sd.frac.dim.index       | tandard deviation of fractal dimension index.                |
| min.frac.dim.index      | the minimum fractal dimension index.                         |
| max.frac.dim.index      | the maximum fractal dimension index.                         |
| total.core.area         | the sum of the core areas of the patches (m2).               |
| prop.landscape.core     | proportional landscape core                                  |
| mean.patch.core.area    | mean patch core area.                                        |
| sd.patch.core.area      | standard deviation of patch core area.                       |
| min.patch.core.area     | the minimum patch core area.                                 |
| max.patch.core.area     | the maximum patch core area.                                 |
| prop.like.adjacencies   | calculated from the adjacency matrix, which shows the frequency with which different pairs of patch types (including like adjacencies between the same patch type) appear side-by-side on the map (measures the degree of aggregation of patch types). |
| aggregation.index       | computed simply as an area-weighted mean class aggregation index, where each class is weighted by its proportional area in the landscape. |
| splitting.index         | based on the cumulative patch area distribution and is interpreted as the effective mesh number, or number of patches with a constant patch size when the landscape is subdivided into S patches, where S is the value of the splitting index. |
| effective.mesh.size     | equals 1 divided by the total landscape area (m2) multiplied by the sum of patch area (m2) squared, summed across all patches in the landscape. |


## Calculation of Fragmentation Indicators

Each of these indicators is calculated for a given polygon based on the underlying
distribution of pixels in a binary forest mask. For this reason, we can easily
use the functionality explained in previous vignettes to calculate fragmentation
statistics. First, let us read in our data and calculate yearly binary forest 
cover maps.

```{r read_data}

# reading data
treeCover = raster(system.file("extdata", "pkgTest_treecover2000.tif", package = "mapme.forest"))
names(treeCover) = "tree.cover"
lossYear = raster(system.file("extdata", "pkgTest_lossyear.tif", package = "mapme.forest"))
names(lossYear) = "loss.year"

# applying preprocessing
treeCover_binary = prepTC(treeCover,
                          thresholdClump = 10,
                          thresholdCover = 50)
treeCover_yearly = getTM(treeCover_binary,
                         lossYear,
                         2000:2018)

# applying minimal mapping unit for the yearly layers
treeCover_yearly = stack(lapply(1:nlayers(treeCover_yearly), function(l){
  treeCover_yearly[[l]] = prepTC(treeCover_yearly[[l]], thresholdClump = 10)
}))

names(treeCover_yearly) = paste("y", 2000:2018, sep = "")

# plotting only first and last layer
plot(treeCover_yearly[[c(1,19)]])

```

In case you have read the [vignette on areal statistics](vignette("area_stats")
you should be already familiar with the above code. If not, please read it before
continuing to read about the fragmentation statistics. 

In the plot above we can already see that our area of interest is characterized
by heavy tree cover losses during the last two decades. We should be able to 
catch this dynamic in our fragmentation statistics as well, since the forest area
becomes smaller and more patchy with the time possibly altering the landscape 
statistics as well. 

To calculate fragmentation statistics we developed a routine called `FragStatsCalc()`.
We can either calculate selected indicators from the table above by specifying them
in a character vector called `FragStats` or we simply calculate all indicators
by specifying `FragStats = "all"` which conveniently also is the default setting.

```{r calc_fragstats}

# read in polygons to extract fragmentation statistics for
aoi = st_read(system.file("extdata", "aoi_polys.gpkg", package = "mapme.forest"))
ncores = 2
if("Windows" %in% Sys.info()['sysname']) ncores = 1

frag_statistics = FragStatsCalc(inputRasterFiles = treeCover_yearly, 
                                latlon = T, 
                                polyName = "id", 
                                studysite = aoi, 
                                FragStats = "all", 
                                ncores = ncores)

cat("Number of rows: ", nrow(frag_statistics))
cat("Number of columns: ",ncol(frag_statistics))
str(frag_statistics[grep("patch.density", names(frag_statistics))])

```

With the above code we calculated all available fragmentation indicators for our
eight polygons. The resulting `frag_statistics` object basically the same `sf`-object
called `aoi` only with its dataframe amended by 792 variables. We printed the structure
of all variables matching the pattern `patch.density` above. As you can see, the 
first three parameters are not associated with a year. These parameters are the
trend values with `Dif` representing the difference between the first and last observation
(between 2000 and 2018 in the present case), `Sig` representing the significance and
`Slope` the slope of the sens-slope-trend calculation (click [here](https://www.rdocumentation.org/packages/trend/versions/1.1.2/topics/sens.slope)
to learn more about the algorithm behind the sens slope). After these parameters,
the values for each specific year are shown. Not that each of the numeric vectors
is of length eight, each representing one of the polygons of the `aoi` object. 


## Non-Spatial Visualization

Now let us plot the dynamic of some selected indicators over time from 2000 to 2018.
We choose the number of patches, the average patch area and the total edge length
for visualization. We will not discuss the trend statistics here, so we will drop
them from out target data frame. Before handing the data to `ggplot2` we need to
reshape the data from a wide representation to a long representation (see [here](http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/)
for a more thorough discussion of reshaping R dataframes).

```{r}

# select target variables
target_variables = st_drop_geometry(frag_statistics[grep(paste("id", "n.patches", "total.edge", "mean.patch.area", sep="|"), names(frag_statistics))])
# drop unneeded variables
target_variables = target_variables[,-grep(paste("Dif", "Sig", "Slope", sep="|"), names(target_variables))]
# reshape and plot with dplyr syntax and pipe operator
plt = target_variables %>% 
  gather("year", "value", -id) %>%
  separate(year, into = c("var","year"), "_") %>%
  mutate(year = as.factor(str_sub(year, -4, -1)),
         id = as.factor(id),
         var = as.factor(var)) %>%
  ggplot(aes(x=year, y = value)) +
  geom_line(aes(color = id, group = id)) +
  facet_wrap(~var, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggplotly(plt)

```

As we can observe from the plot above, the average patch area is significantly reduced
over the years. This is an indicator that the landscape which consisted of a few
very large patches became more and more patchy with smaller average areas. This 
is also observed in the second parameter, the total number of patches. Up until the year
2015 or so, the number of patches increases in almost all polygons from below 10
up to 20 or even above 30. However, the number of patches decreases during the last
years which might point towards the complete disappearance of some forest patches 
towards the end of the observation period. The total length of edges shows a very 
similar pattern with a steep increase during the first years but an even steeper 
decrease towards the end of the time series. This too, indicates that a high number
of forest patches are completely disappearing. 

With this we have demonstrated the calculation and interpretation of some of 
fragmentation parameters on a polygons basis. Which parameters are important 
for a certain question, however, remains to be discussed in a specific project 
context.

## Spatial Visualization

As you might have already guessed, the data we obtain by this calculation is hardly
suitable for a representation on a map. Of course we could visualize some of
the parameters by using a cholorphlet map-style like this:

```{r}
tmap_mode("view")

tm_basemap("OpenStreetMap") +
  tm_shape(frag_statistics) +
  tm_polygons("total.core.area_y2000")
```

However, we have a very high number of variables and the spatial information on 
the variables is not very explicit in a map representation like above.
For that reason, we developed an additional workflow which allows you to
display the fragmentation statistics in a raster. Of course, this does not come for
free but it is associated with several disadvantages and simplifications which 
are going to discuss now.

The idea behind the visualization as a raster is to treat different zones of the raster
as polygons for which to retrieve the fragmentation statistics. When we align
theses zones in an ordered fashion we actually create a new raster with a coarser
resolution than the original one. The values in the new raster actually represent 
what is going on in the corresponding cells of the original raster. 
However, the values will vary based on the decision how we align these zones and
how big they are, meaning which resolution we choose for the new raster. 

Let us start by creating a regular spaced grid for our study area. Because
we will have to specify a resolution for our new raster set it might be a good
idea to project our data first to a metric CRS. We are based in Vietnam, so
UTM zone 48N seems reasonable.

```{r}
utm = "+proj=utm +zone=48 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
treeCover_proj = projectRaster(treeCover_yearly, crs = CRS(utm), method = "ngb")
res(treeCover_proj)
```

This shows us, that after projecting our raster dataset we have a resolution of 
`r res(treeCover_proj)[1]` meters in the x direction and a resolution of `r res(treeCover_proj)[2]`
meters in the y direction. Now, we need a grid which`s cell sizes aggregates our
original raster by a certain factor. Currently one pixel comprises an read of
`r round(res(treeCover_proj)[1] * res(treeCover_proj)[2])` m². Let us say, we want
to aggregate our raster by a factor of 200, that means that our target cells should
cover an area of `r round((((res(treeCover_proj)[1] * res(treeCover_proj)[2])) * 200) / 10000, 2)` ha.
We will use that number to calculate how many cells of the original resolution 
are needed to get one target cell, which is simply put the  square root of the 
aggregation factor.
From there, we can calculate how many columns and rows we would get when we simply
aggregated the original cells to the coarser resolution. We finally create a matrix
by multiplying the number of cells of our original resolution within one unit of
our target resolution with the number of columns and rows of our desired target resolution.



```{r}
agg_factor = 200
n_cells = round(sqrt(agg_factor))
ncols = ceiling(ncol(treeCover_proj) / n_cells)
nrows = ceiling(nrow(treeCover_proj) / n_cells)
m = matrix(nrow = n_cells * nrows, ncol = n_cells * ncols)

cat("Dimension of our target matrix is: ", dim(m))
cat("Dimension of our original raster is: ", dim(treeCover_proj))
```
We can see that our target matrix is slightly bigger than our original raster.
This is because we used the `round()` and `ceiling()` function to obtain even numbers
which are needed to create a matrix. We will now fill our created matrix in a nested
for loop, assigning the same values to cells which comprise one cell in our target
resolution. What we need for this is simply the variable `n_cells` representing
the square root of our aggregation factor.

```{r}

for (i in 1:ncols){
  for (j in 1:nrows){
    m[(n_cells*j - n_cells + 1) : ((n_cells*(j+1) - n_cells)) , (n_cells*i - n_cells + 1) : (n_cells*(i+1) - n_cells)] = 1 * i * j
  }
}

```

Now we have got a matrix which resembles our original raster and assigning a common
value to all cells which comprise one cell in our target resolutions. These are 
the zones we are going to calculate the fragmentation indicators for. 
There remains one small problem, however, which is that our target matrix currently
is slightly larger than our original raster. To account for this, we will simply 
delete the rows and columns which are to many evenely on all sights fo the matrix.
However, that means that we reduce the size of the zones on the edges of the matrix
compared to the inner zones. This makes the outer zones incomparable to the inner
ones. In practice, this means that we should always calculate the indicators with 
a raster slightly bigger than our area of interest and crop it in the end. 

```{r}

diff_nrow =  nrow(m) - nrow(treeCover_proj)
diff_ncol =  ncol(m) - ncol(treeCover_proj)
m = m[(((diff_nrow / 2) ) +1): (nrow(m) - (diff_nrow / 2)) , ((diff_ncol / 2) +1): (ncol(m) - (diff_ncol / 2)) ]
cat("Dimension of our target matrix is: ", dim(m))
cat("Dimension of our original raster is: ", dim(treeCover_proj))

```
Now, we got a matrix which exactly resembles our original raster and the values in
the matrix indicate the zones of our target resolution. What we need now, is
to transform this information to an sf-object to be able to calculate the 
fragmentation statistics as we know it. For this to work, we first have to transform
our matrix to a raster. In addition to the `raster` package we will use some functions
from `stars` and `sf`, as these work more efficiently when polygonizing raster values.

```{r}
grid = raster(m)
crs(grid) = crs(treeCover_proj)
extent(grid) = extent(treeCover_proj)
grid_stars = stars::st_as_stars(grid)
polys = st_as_sf(grid_stars[1], as_points = FALSE, merge = T)
polys$id = 1:nrow(polys)
plot(polys["id"])

```
We will now calculate the fragmentation statistics for each of these polygons
just in the same way as we saw it before. Because this command is executed
each time the markdown is rendered we will try to speed up the computation time
by only calculating one indicator for only one of the layers in the `treeCover_proj` 
object.

```{r gridded_stats, cache=TRUE}
grid_fragstats = FragStatsCalc(inputRasterFiles = treeCover_proj[[1]],
                               latlon = F,
                               polyName = "id",
                               studysite = polys,
                               FragStats = "patch.density",
                               ncores = ncores)
results = st_drop_geometry(grid_fragstats)
result_raster = raster(matrix(data = results$patch.density_V1, ncol = ncols, nrow = nrows))
result_raster[] = NA
crs(result_raster) = crs(treeCover_proj)
extent(result_raster) = extent(treeCover_proj)
result_raster = setValues(result_raster, as.matrix(results$patch.density_V1))
plot(result_raster)

```

Now, all that is left is to crop the raster to a smaller extent in order to get rid
of the cells on the edge which do not have meaningful values.

```{r}
bbox = st_as_sfc(st_bbox(result_raster))
bbox_buffer = st_buffer(bbox, dist = -500)
cat("Uncropped raster with buffered extent")
plot(result_raster)
plot(bbox_buffer, add = T)
results_cropped = crop(result_raster, st_as_sf(bbox_buffer))
cat("Cropped raster")
plot(results_cropped)
```


